The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2011-03-09 smurf_paper.tex
overview of map-making and draft of FLT and AST

\subsubsection{Principal component analysis}

@@

There are two general families of existing map-making strategies that
were considered. The first are \emph{maximum likelihood} methods that
were originally pioneered for the analysis of Cosmic Microwave
Background (CMB) maps. Here, the pixels in the map are considered to
be free parameters in a model-fitting problem, and the data are the
raw time-series. By estimating the full noise covariance matrix for
the data, it is then possible to estimate the maximum-likelihood
map. While direct inversion solutions to this problem are untenable,
some recent efforts to optimize these algorithms have enabled the
reduction of smaller data sets in a reasonable amount of time. While
maximum likelihood maps are, theoretically, the best maps that can be
made from a noise point of view, the required computing resources and
execution time would not have been feasible for the proposed real-time
data reduction and data archiving model.

@@

As for the noise, $\mathbf{n}_i(t)$, we break it down into several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component.

The primary goal of map-making is to model and remove
$\mathbf{n}^\mathrm{f}_i$ and $\mathbf{n}^\mathrm{c}$ from the
time-series. By re-gridding the remaining data we can then hope to
approach the theoretical noise limit in the map.

@@

@@

Give details about each model component, roughly in the order they are
calculated.

\subsection{Common-mode estimation}

@@

We estimate the common-mode, remove it from the time series, use it to
flag bad bits of data, check for time-varying bolometer
responsivities... what doesn't it do?!

\subsection{Iterative filtering}

The FFT filter can do high/low-pass filtering. We can also do notches
and whitening.

@@

\subsection{Noise estimation}

Measure a white noise level from the power spectrum for each detector
to use as a weight.

\subsection{Map estimation}

In the map, use weighted standard error to estimate noise. This is
accurate for a single pixel, but there is still spatially correlated
noise in the maps which we can show with some figures.

@@

>>>>

\subsubsection{Principal component analysis}
\label{sec:pca}

@@

There are two general families of existing map-making strategies that
were considered. The first are \emph{maximum likelihood} methods that
were originally pioneered for the analysis of Cosmic Microwave
Background (CMB) maps. Here, the pixels in the map are considered to
be free parameters in a model-fitting problem, and the data are the
raw time-series. By estimating the full noise covariance matrix for
the data, it is then possible to estimate the maximum-likelihood
map. While direct inversion solutions to this problem are untenable,
some recent efforts to optimize these algorithms have enabled the
reduction of smaller data sets in a reasonable amount of time
\citep[e.g.][]{patanchon2008}. While maximum likelihood maps are,
theoretically, the best maps that can be made from a noise point of
view, the required computing resources and execution time would not
have been feasible for the proposed real-time data reduction and data
archiving model.

@@

As for the noise, $\mathbf{n}_i(t)$, we break it down into several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) + \mathbf{n}^\mathrm{f}_i(t) +
  \mathbf{n}^\mathrm{c}(t),
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white (thermal)
noise, $\mathbf{n}^\mathrm{f}_i(t)$ is low-frequency noise that is
\emph{not} correlated from bolometer-to-bolometer, and
$\mathbf{n}^\mathrm{c}(t)$ is a (predominantly low-frequency)
correlated or \emph{common-mode} component. The primary goal of
map-making is to model and remove $\mathbf{n}^\mathrm{f}_i$ and
$\mathbf{n}^\mathrm{c}$ from the time-series. By re-gridding the
remaining data we can then hope to approach the theoretical noise
limit in the map. This procedure is summarized in Fig.~\ref{fig:dimm}.

@@

First, the data files are read from disk, concatenated into contiguous
arrays in memory, and run through a pre-processing stage that performs
several operations including spike and sudden level step removal to
enforce continuity in the time-series. Then, the iterative process
within the dashed box begins. It is generally assumed that the
dominant noise component is the common-mode, which is presumably
dominated by large variations in the fridge temperature and sky noise
(as argued in Section~\ref{sec:pca}). This signal is estimated as the
average signal observed by all of the bolometers at each time step,
and then removed from all of the bolometers. Once this large signal
has been removed, we apply the extinction correction, which is
determined completely from external measurements. Next, residual noise
that differse from bolometer-to-bolometer (primarily low-frequency) is
removed using an FFT-based filter. The resulting ``clean'' time-series
are then re-gridded to produce a map estimate using nearest-neighbour
sampling. Since each map pixel averages many bolometer samples
together, it is fundamentally much less noisy than the bolometer
time-series. We can therefore project the map back into the time
domain, and create relatively noiseless estimates of what each
bolometer would have seen in the absence of noise sources, and remove
that from the time-series. This final step in the iteration gives a
residual signal (with low-frequency and astronomical noise components
removed) that may be used to measured the white noise level in each
detector, and also to estimate an approximate $\chi^2$ for the model
to track convergence. As we will discuss in the next sections, each
signal component estimate is biased slightly from the presence of
signal from other components, so the entire process is
iterated. However, before the next iteration begins, we first undo the
multiplicative factors introduced by the extinction correction, and
also relative gains that may optionally be estimated in the
common-mode estimation phase.

In practice, the map-maker is highly configurable, both in terms of
the order in which model components are fit (and whether they are fit
at all), and the details of \emph{how} they are fit. In the following
sections we describe these details.

@@

In addition to specifying when each model component is calculated
(normally an attempt is made to order them by decreasing amplitude in
the time-series), there are a number of parameters that control their
operation.

\subsection{\texttt{COM,GAI}: common-mode estimation}

@@

We estimate the common-mode, remove it from the time series, use it to
flag bad bits of data, check for time-varying bolometer
responsivities... what doesn't it do?! Also mention the optional
presence of GAI.

\subsection{\texttt{FLT}: fourier transform filter}

This model takes the FFT of the bolometer data, and can apply both
high- and low-pass filters, as well as notch filteres, at hard
frequency edges specified by the user. Alternatively, the frequency
edges of the filters may be defined in terms of an \emph{angular
  scale}, but converted into a frequency through knowledge of the mean
telescope slew speed. The time-series may optionally have apodization
applied before the transform to avoid ringing (primarily caused by
wrap-around discontinuities at the ends of the time-series). The
signal that is \emph{removed} from the time-series by this process are
stored in the \texttt{FLT} container array.

\subsection{\texttt{SMO}: time-domain smoothing filter}

@@

\subsection{\texttt{AST}: map estimation}

Map estimation is accomplished using a simple nearest-neighbour
resampling of the data onto a pre-defined map grid. For the $i$th map
pixel $\mathbf{m}(x_i,y_i)$, the brightness is estimated as the
weighted average of the bolometer data samples $\mathbf{b}_j$ that
land within that pixel (from any bolometer or point in time),
%
\begin{equation}
  \mathbf{m}(x_i,y_i) = \frac{\sum_j \mathbf{w}_j \mathbf{b}_j }
                             { \sum_j \mathbf{w}_j } .
\end{equation}
%
For the initial iteration the weights $\mathbf{w}_j$ are set to 1, but
subsequently they are set to $1/\sigma_j^2$, the estimated inverse
variance expected from the bolometer white noise levels as discussed
in Section~\ref{sec:noi}. This weighting scheme is sensible provided
that the bolometer data have no correlated (i.e., low-frequency)
noise.

In addition, a variance map $\mathbf{v}(x_j,y_j)$ is estimated. The
default behaviour is to estimate this weighted error on the mean given
the scatter in the weighted samples. This is accomplished by dividing
the \emph{biased weighted sample variance} by the number of samples
that went into the average (akin to the formula for standard error on
the mean, but accounting for weights),
%
\begin{equation}
\mathbf{v}(x_j,y_j) = \frac{\sum_j \mathbf{w}_j
                            \sum_j \mathbf{w}_j \mathbf{b}_j -
                            \left( \sum_j \mathbf{w}_j \mathbf{b}_j \right)^2 }
                           { N \left( \sum_j \mathbf{w}_j \right)^2 },
\end{equation}
%
where $N_j$ is the total number of bolometer samples that land in the
pixel. We decided not to use the \emph{unbiased} estimator since in
practice it would require accumulating an additional array of values
at every map pixel, and only results in a small difference where there
are less than $\sim10$ samples per pixel (a situation that is almost
never encountered in a SCUBA-2 map except in the edge pixels).

Finally, once the map estimation is complete, the map is projected
into the time domain (the signal that would be produced in each
bolometer by the signal represented by the map) and removed.

In addition, the \texttt{AST} model can use the map and its variance
to provide additional despiking of the time-series. Unlike the
time-domain despiker described ???, this calculation uses the scatter
in the population of samples used to estimate the map pixel value
(from different times and bolometers) to reject outliers. This method
is more robust against false-positive detections of bright/compact
astronomical sources since transient features in the time-series are
unlikely to occur by chance whenever a bolometer crosses a specific
location on the sky, whereas real astronomical sources are at fixed
spatial locations.

The estimated variance, $\mathbf{v_p}(x_i,y_i)$, of the normalized
weighted samples that land in the $i$th map pixel is simply the biased
weighted sample variance (i.e., the variance map value multiplied by
the number of samples)
%
\begin{equation}
  \mathbf{v_p}(x_i,y_i) = N_i \mathbf{v}(x_i,y_i).
\end{equation}

In order to compare the weighted \emph{differences} between the
samples and the map values, [$\mathbf{b}_j - \mathbf{m}(x_i,y_i)$] to
$\mathbf{v_p}$, they must be scaled appropriately. We define a
normalized difference, $\mathbf{d}_j$, in such a way that the variance
of this new variable gives the weighted sample variance of the
underlying data points,
%
\begin{eqnarray}
  \frac{\sum_j \mathbf{d}_j^2}{N} &=&
  \frac{\sum_j \mathbf{w}_j [\mathbf{b}_j - \mathbf{m}(x_i,y_i)]^2}
       {\sum_k \mathbf{w}_k} \\
   \Rightarrow \mathbf{d}_j^2 &=& \frac{N \mathbf{w}_j [\mathbf{b}_j -
       \mathbf{m}(x_i,y_i)]^2}{ \sum_k \mathbf{w}_k} .
\end{eqnarray}
%
The map-based despiker flags those $\mathbf{d}_j$ that are further
than some threshold number of standard deviations,
$\sqrt{\mathbf{v_p}}$, away from zero so that they are not used in
subsequent iterations.

\subsection{\texttt{NOI}: noise estimation}
\label{sec:noi}

The primary purpose of the noise component, \texttt{NOI} is to measure
the noise in each bolometer residual once all of the other models have
been fit and removed. This noise may then be used to estimate weights
for the data in subsequent iterations.

@@

\subsubsection{Down-sampling}

\subsubsection{Whitening filter}
